Modelling last glacial cycle ice dynamics in the Alps

### 2017-09-20 articles/cycle/alps-cycle.tex
Rephrase palaeo-climate section.
I just realised that an important point about LGM timing is missing.

    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice cap, showing that alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences and
    self-sustained ice domes. Finally, the Last
    Glacial Maximum advance, often considered synchronous, is here modelled as
    a time-transgressive event, with some glacier lobes reaching their maximum
    as early as 27\,ka, and some as late as 21\,ka. Modelled ice thickness is
    about 900\,m higher than observed trimline elevations, yet our simulation
    predicts little erosion at high elevation due to cold ice conditions.

@@

    For the three palaeo-temperature records and the two palaeo-precipitation
    parametrizations used, the model yield significant ice volume build-up
    during MIS~4 and 2 (Fig.~\ref{fig:timeseries}). All simulations also
    yield important glaciations during MIS~5 and 3, but timing and amplitude
    varies a lot between the different forcing. All simulations overestimate
    ice cover during the Younger Dryas. This might be due partly to the 2\,km
    resolution, except for GRIP which largely overestimates Younger Dryas ice
    cover.

    Even more importantly, all simulations yield very strong ice volume
    variability which is the result of multiple glacier advance and retreats
    over the foreland, many more than two. Palaeo-precipitation reductions
    partly results in a smoother modelled ice volume time series and less
    extreme variations, but variability remains high. Ice volume variability
    is the smallest for the EPICA palaeo-temperature record, which has the
    least temperature variability.

@@

    To select an optimal climate forcing, we look at the best known alpine
    glaciations during MIS~2 and 4. During MIS~2, all forcing yield ice extent
    of the same order of magnitude (Fig.~\ref{fig:footprints}). This is
    because the temperature records have been scaled to resul in similar
    ice covered area for the Rhine Glacier piedmont lobe. Looking at ice cover
    on other parts, all simulations tend to overestimate ice extent in the
    eastern part of the model domain, and to underestimate it in the western
    part, as compared to the reconstructed LGM margin. Regarding the western
    part, there are discussions though, as if the Lyon Lobe actually dates from
    MIS~2 or from an older glaciation. Regarding the eastern part, the MIS~2
    glaciation has been mapped in detail and dated (?), so that there is no
    doubt that our model overestimate ice cover in this region. This could
    indicate that during LGM the east-west precipitation gradient over the Alps
    was lower than today, or that temperature depression in the east was lower
    than in the west.

    There are spatial differences though: MD01-2444, and to a greater extent
    GRIP, forcings, tend to overestimate ice cover in all peripherical ranges:
    the Vosges, Black Forest, Bavarian Forest, Dinaric Alps and Pennine. This
    is because these records have a higher temperature variability so that
    after scaling they comport brief periods of cold climate, not long enough
    to develop the alpine ice cap, but long enough to build up ice on the
    peripherical ranges.

    The extent of ice cover during MIS~4 shows more sensitivity to the choice
    of palaeo-climate forcing. Using the GRIP and MD01-2444, glaciers extend
    well beyound the reach of documented moraines and glacial erratic terrain.
    Because this extent corresponds to a short-term cold climate event,
    palaeo-precipitation reductions help to greatly reduce this excessive
    modelled ice cover, yet in both cases, modelled ice extent and volume
    during MIS~4 remains higher than during MIS~2. EPICA forcing, on the other
    hand, yield a MIS~4 glaciation that is only slightly less expensive than
    the MIS~2, with only little sensitivity of the glaciated area to
    palaeo-precipitation reductions.

    Based on the above consideration on timing of the LGM ice extent and
    ice extent during MIS~4, we chose EPICA as our optimal temperature record
    for the rest of this paper. As a conservative approach in regard to quick
    ice volume fluctuations, we choose to include palaeo-precipitation
    correction in the following 1\,km-resolution simulation.

    \note{An important assumption in the method is that the MIS~4 glaciation
          was less (or equally) extensive than the MIS~2, i.e. that the LGM
          outline from \citet{Ehlers.etal.2011} dates indeed from the LGM
          (MIS~2). A different assumption could be that this outline represents
          the most extensive glaciation in the last 120\,ka. If we allow the
          MIS~2 glaciation to be a bit smaller than MIS~4, simulations driven
          by GRIP and MD01-2444 also give reasonable results.}

@@

    In addition, our model results depict a polythermal LGM Alpine ice cover.
    Due to sub-freezing temperatures applied in the climate forcing of the
    model, the entire Alpine ice field is capped by an upper layer of cold ice.
    In major Alpine valleys, such as the upper Rhone Valley, important ice
    thickness and strain heating contribute to form a layer of temperate ice
    near the glacier base, allowing basal melt, sliding and potentially
    erosion. On the other hand, on the highest mountains ice cover is to thin
    and too static to form temperate ice, resulting in cold ice down to the
    bed, preventing potential erosion.

>>>>

    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences and
    self-sustained ice domes. Finally, the Last
    Glacial Maximum advance, often considered synchronous, is here modelled as
    a time-transgressive event, with some glacier lobes reaching their maximum
    as early as 27\,ka, and some as late as 21\,ka. Modelled ice thickness is
    about 900\,m higher than observed trimline elevations, yet our simulation
    predicts little erosion at high elevation due to cold ice conditions.

@@

    For the three palaeo-temperature records and the two palaeo-precipitation
    parametrizations used, the model yield significant ice volume build-up
    during MIS~4 and 2 (Fig.~\ref{fig:timeseries}b), corresponding to documented
    glaciation periods in the Alps \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
    All simulations also yield important glaciations during MIS~5 and 3, but
    their timing and amplitude varies significantly depending on the climate
    forcing used. All six simulations overestimate
    ice cover during the Younger Dryas \citep[cf. e.g.,][]{Ivy-Ochs.etal.2009}.
    This is partly because the 2\,km resolution is too coarse to resolve
    Younger Dryas Alpine glaciers, but the large total ice volume
    overestimation resulting from the GRIP temperature forcing
    (Fig.~\ref{fig:timeseries}b, blue curves) can not be explained by
    resolution alone.

    Even more importantly, all simulations yield very strong ice volume
    variability, indicative of many more than two or three cycles of glacier
    advance and retreats onto the foreland (Fig.~\ref{fig:timeseries}b).
    Palaeo-precipitation reductions dampen some of the small-scale variability,
    but they have little effect on the millennial scale that characterise those
    cycles (Fig.~\ref{fig:timeseries}b, light colour curves). Finally, the
    EPICA palaeo-temperature record has least variability the among records used
    (Fig.~\ref{fig:timeseries}a, red curves). This results in fewer variability
    in the total ice volume, and more restrictive glaciations during MIS~5
    and~3, (Fig.~\ref{fig:timeseries}b, red curves).

    \todo{comment on timing of the LGM}

@@

    During MIS~2, all six simulations yield comparative cumulative glaciated
    areas (Fig.~\ref{fig:footprints}a). This result is inherent to the
    palaeo-climate forcing approach used, which involved a linear scaling of
    each palaeo-temperature anomaly record to a geomorphological
    reconstruction of the area glaciated by the Rhine Glacier piedmont lobe
    (Sect.~\ref{sec:paltemp}, Fig.~\ref{fig:footprints}a, black rectangle).
    Concerning other areas, all simulations underestimate glaciated area in the
    western part of the model domain, and overestimate it in the western
    part, relative to the geomorphological ice limits
    (Fig.~\ref{fig:footprints}a). This result might indicate that the LGM
    temperature depression, here considered homogeneous, was actually lower in
    the Eastern Alps and in the Western Alps. Alternatively, the east-west
    gradient in precipitation existing today (Fig.~\ref{fig:inputs}d) was
    perhaps enhanced during the LGM, as indicated by pollen reconstruction
    \citep{Wu.etal.2007} and positive degree-day modelling of central European
    palaeo-glaciers \citep{Heyman.etal.2013}.

    Beside this general pattern, there exist differences in glaciated area
    depending on the palaeo-climate forcing used. The MD01-2444, and to a
    greater extent, the GRIP, palaeo-temperature records, tend to overestimate
    MIS~2 ice cover on all peripheral ranges that surround the Alps
    (Fig.~\ref{fig:footprints}a and c). This is particularly true when
    precipitation corrections are applied (bright colours). In fact, both
    palaeo-temperature records have a higher temperature variability and
    contain brief spells of cold temperatures, which are too short to develop a
    fully grown Alpine ice sheet, but long enough to build up ice cover on a
    smaller scale on these peripheral ranges (Fig.~\ref{fig:timeseries}). On
    the other hand, peripheral glaciation modelled in the EPICA simulation
    (Fig.~\ref{fig:footprints}b) is in relatively good agreement with the
    geomorphological reconstructions.

    The modelled extent of glaciation during MIS~4 depicts a more pronounced
    sensitivity to the choice of palaeo-climate forcing used. Using the GRIP
    and MD01-2444 palaeo-temperature records, Alpine glaciers are modelled to
    extend well beyond the reach of documented LGM end-moraines and glacial
    erratic terrain (Fig.~\ref{fig:footprints}d and f). Because, in both cases,
    the modelled glaciation extent corresponds to brief cold spells in the
    palaeo-temperature forcing (Fig.~\ref{fig:timeseries}a),
    palaeo-precipitation reductions greatly reduce the excessive modelled ice
    cover (Figs.~\ref{fig:timeseries}b and \ref{fig:footprints}d and f, light
    colours). However, with or without palaeo-precipitation corrections,
    modelled ice extents (Fig.~\ref{fig:footprints}d and f) and volumes
    (Fig.~\ref{fig:timeseries}b) are modelled to be higher during MIS~4 than
    during MIS~2. This is in contradiction with geological evidence
    \citep{Preusser.2004, Ivy-Ochs.etal.2008}.

    The EPICA palaeo-temperature forcing, on the other hand, yields a MIS~4
    glaciation that is only slightly less expensive than the MIS~2 glaciation
    (Fig.~\ref{fig:footprints}e). The sensitivity of the glaciated area to
    palaeo-precipitation reductions is mostly limited to the southern terminal
    lobes of central-Alpine glaciers, which are slightly less extensive during
    MIS~2 and 4 (Fig.~\ref{fig:footprints}b and e).

    Based on the above considerations on timing of the LGM ice extent and ice
    extent during MIS~4, we chose EPICA as our optimal palaeo-temperature
    record for a more detailed and higher-resolution comparison of modelled
    glacier dynamics to available geological evidence
    (Sect.~\ref{sec:results}). As a conservative approach in regard to the
    rapid ice volume fluctuations, we choose to include palaeo-precipitation
    correction in the following comparison.

@@

    In addition, our model results depict a polythermal LGM Alpine ice cover.
    Due to sub-freezing temperatures applied in the climate forcing of the
    model, the entire Alpine ice sheet is capped by an upper layer of cold ice.
    In major Alpine valleys, such as the upper Rhone Valley, important ice
    thickness and strain heating contribute to form a layer of temperate ice
    near the glacier base, allowing basal melt, sliding and potentially
    erosion. On the other hand, on the highest mountains ice cover is to thin
    and too static to form temperate ice, resulting in cold ice down to the
    bed, preventing potential erosion.
